---
editor: source
---

# Good coding practices

In the last chapters, you have learned how to make your data and code findable, accessible, interoperable, and reusable.
In this chapter, you will learn good coding practices to help other researchers and future you with the code you used for your analysis.

## Code Style {#sec-codestyle}

One important aspect that fosters understandability of code is the code style.
To make this more clear, let's think of an example.
Consider yourself in a research project where you want to examine the robustness of the finding that women show higher neuroticism values than men.
You have found a lot of papers examining this relationship, usually finding that indeed women show higher values than men.
Now you are sitting in front of a paper from Wittkuhn et al. (2024).
You want to check if the convenient finding of the authors is due to an error in the code of actually correct.
Compare the following two examples of code.
Which one would you prefer when you want to rerun analyses and why?

```{r}
#| label: bad-code-example
#| eval: FALSE
#| code-line-numbers: true

library(package=tidyverse)
library(psych)#load packages
#get data
DATA<-bfi
#get means of neuroticism-items by gender
DATA$N <- rowMeans(DATA[16:20], na.rm=TRUE)
describeBy(DATA$N, DATA$gender)

#Result: women have higher neuroticism values then men (descriptively)
t.test(N ~gender, data, alternative= "l")#confirmation
#Plot Result in Boxplot
DATA$gender <- factor(DATA$gender, levels= c(1,2), labels =c("male","female"))
ggplot(DATA)+
  geom_boxplot(aes(gender, N, group = gender, fill = gender),show.legend = F)+
  scale_fill_manual(values = c("lightblue", "pink"))
#Done
```

```{r}
#| label: good-code-example
#| eval: false
#| code-line-numbers: true

## Investigating the relationship of neuroticism and gender ##
## Hypothesis: Women show higher values on a neuroticism-scale than men ##

# load required packages
library(psych) #for statistical computations, also contains data
library(tidyverse) #for data wrangling and visualization

# get dataset
data <- bfi #Big Five Inventory

# familiarize with dataset
str(data)
head(data)
print(bfi.dictionary) #getting itemformulations
?bfi #more background information about the data

# create scale variable that contains the means of all N-items of each person
data$N <- rowMeans(
  data[16:20], #select corresponding Neuroticism-Items
  na.rm=TRUE #remove missing values
  )

# get labels for gender values
print(bfi.dictionary) #males = 1, females = 2
data$gender <- factor(
  data$gender,
  levels = c(1, 2),
  labels = c("male", "female")
)

# check correct labels
str(data$gender)

# get descriptive differences
describeBy(
  x = data$N,
  group = data$gender #group values by gender
)
      # Result:
        # Males:    M = 2.95, SD = 1.14
        # Females:  M = 3.27, SD = 1.21
      # Descriptive Statistics congruent to hypothesis.

# check for significant differences
t.test(
  N ~ gender,
  data = data,
  alternative = "less" #assume that men show lower values than women
)
      # Result: Women show significantly higher values in neuroticism-scale than men. 
        # t = 6.727, p < .001
      # Significance testing also congruent to hypothesis

# plot results as a boxplot
ggplot(
  data = data,
  mapping = aes(
    x = gender,
    y = N
    )
  )+
  geom_boxplot(
    mapping = aes(
      group = gender, #two separate boxplots for each gender
      fill = gender #two separate colors for each gender
    ), 
    show.legend = FALSE
  )+
  scale_fill_manual( #set up different colors manually
    values = c(
      "lightblue",
      "pink"
    )
  )
```

## Tidy Data

As you might have noticed when you worked through the @sec-codestyle section, the dataset `bfi` was organized well.
That does not always hold true, when you work with datasets.
In fact, most of the time you are facing datasets, you will spend more time organizing your data than running real analysis.
A helpful concept is **tidy data**, which is a common guideline for datasets to be organized.
Following tidy data guidelines will help you running analyses and getting the most out of your data.

::: callout-important
## Tidy Data

Tidy datasets follow three basic rules:

1.  Each variable is a column; each column is a variable.
2.  Each observation is a row; each row is an observation.
3.  Each value is a cell; each cell is a value.
:::

![Visualization of Tidy Data by Wickham et al. ([2024](https://r4ds.hadley.nz/data-tidy)). Used under a [CC BY-NC-ND 3.0 US license](http://creativecommons.org/licenses/by-nc-nd/3.0/us/).](/images/tidy-1.png){fig-align="center"}

# Acknowledgements & further reading

We would like to express our gratitude to the following resources, which have been essential in shaping this chapter.
We recommend these references for further reading:

| Resources |
|------------------------------------------------------------------------|
| Buhr, J. (2023). *Introduction to Data Analysis with R.* <https://jmbuhr.de/dataintro/> |
| Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2024). *R for Data Science (2e)*. <https://r4ds.hadley.nz> |
|  |
