---
title: "Naming things"
engine: knitr
execute:
  eval: false
  warning: false
  message: false
categories: [beginner, basics]
abstract: |
  In this chapter you will learn how to name things.
---

## Examples

### Bad names

```{zsh filename="Output"}
#| code-copy: false
myabstract.docx
Joe’s Filenames Use Spaces and Punctuation.xlsx
figure 1.png
fig 2.png
JW7d^(2sl@deletethisandyourcareerisoverWx2*.txt
```

### Better names

```{zsh filename="Output"}
#| code-copy: false
2014-06-08_abstract-for-sla.docx
joes-filenames-are-getting-better.xlsx
fig01_scatterplot-talk-length-vs-interest.png
fig02_histogram-talk-attendance.png
1986-01-28_raw-data-from-challenger-o-rings.txt
```

::: {.callout-caution title="Exercise: What is bad or good about these file names?" collapse="false"}
Pause for a moment an think about what exactly makes the file names above bad or better?
Discuss with a learning partner.
:::

## Three principles for (file) names

1. [Machine readable](#machine-readable)
1. [Human readable]((#human-readable))
1. Plays well with default ordering

## Machine readable

A file is machine-readable if information is formatted in a way that a computer can easily read and understand it without needing a human to interpret or manually enter the data.
This often involves using standardized data formats like CSV, XML, or JSON, which are structured so that computer programs can process the information quickly and accurately.

### Regular expression and globbing friendly

A **regular expression**, often called a "regex", is a sequence of characters that defines a search pattern.
It's used to find or match specific combinations of characters within text.
You can think of it as a tool for searching text in a very precise and flexible way.

**Globbing** is a way to use special characters (called "wildcards") to match multiple files or directories in command-line interfaces or scripts.
It helps you select groups of files without typing every single name.

Common wildcards in globbing:

- `*` (asterisk): Matches any number of characters.
For example, `*.txt` matches all files ending with `.txt`.
- `?` (question mark): Matches exactly one character.
For example, `file?.txt` matches `file1.txt` and `file2.txt`, but not `file10.txt`.

So, if you wanted to list all text files in a directory, you could use the pattern `*.txt`, and it would match all files with the `.txt` extension.

Both regular expression and globbing are concepts that help make working with files and data more efficient for automation and scripting tasks.

To make file names regex and globing friendly, avoid:

- spaces
- punctuation
- accented characters
- case sensitivity

::: {.callout-tip title="Find out more: What are regular expressions?" collapse="true"}

A regular expression is a special sequence of characters that helps you to search for patterns in text.
It allows you to find specific words or phrases within a larger body of text.
For example, imagine you have a text document and you want to find all email addresses mentioned in it.
You can use the following regular expression to search for email addresses:

```{zsh filename="Code"}
\b[A-Za-z0-9._%+-]+@[A-Za-z0-9.-]+\.[A-Z|a-z]{2,}\b
```

This pattern will match any text that looks like an email address, such as `example@email.com`.
Imagine that you have a text like the following:

> Please contact me at john.doe@email.com for any inquiries.
You can also reach out to jane.smith@example.com for more information.

You can see regular expression in action with this example:

```{zsh filename="Code"}
echo "Please contact me at john.doe@email.com for any inquiries. You can also reach out to jane.smith@example.com for more information." | grep -Eo "\b[A-Za-z0-9._%+-]+@[A-Za-z0-9.-]+\.[A-Z|a-z]{2,}\b"
```

Here is what happens with this command:

- `echo` is used to display the sample text.
- The pipe (`|`) sends the output of `echo` as input to `grep`.
- `grep -Eo` enables extended regular expressions and prints only the matching part of the line.
- The regular expression `\b[A-Za-z0-9._%+-]+@[A-Za-z0-9.-]+\.[A-Z|a-z]{2,}\b` is used to identify email addresses in the text output by `echo`.

You can also paste the example text into a text file, for example called `example.txt` and then apply the same command to the text file:

```{zsh filename="Code"}
grep -Eo "\b[A-Za-z0-9._%+-]+@[A-Za-z0-9.-]+\.[A-Z|a-z]{2,}\b" example.txt
```

:::

::: {.callout-tip title="Find out more: What is globbing?" collapse="true"}

Globbing refers to the use of wildcard characters (like `*` and `?`) to match multiple files or directories in command-line interfaces and scripting.
Being globbing friendly means file names or paths are structured in a way that makes them easy to match using these wildcards.

For example, let's assume that you have two files like `report_2023_10_01.txt` and `report_2023_10_02.txt`.
You can use a glob pattern like `report_2023_10_*.txt` to match all of them.

:::

#### Example

Excerpt of complete file listing:

```{zsh filename="Output"}
#| code-copy: false
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H01.csv
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H02.csv
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H03.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A01.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A02.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A03.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A04.csv
```

::: {.panel-tabset}

## {{< fa terminal >}} Bash

```{zsh filename="Code"}
ls *Plasmid*
```

```{zsh filename="Output"}
#| code-copy: false
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H01.csv
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H02.csv
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H03.csv
```

## {{< fa brands apple >}} macOS

Put image of macOS screenshot here

## {{< fa brands r-project >}}

```{zsh filename="Code"}
list.files(pattern = "Plasmid") |> head
```

```{r filename="Output"}
#| code-copy: false
[1] "2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_A01.csv"
[2] "2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_A02.csv"
[3] "2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_A03.csv"
[4] "2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_B01.csv"
[5] "2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_B02.csv"
[6] "2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_B03.csv
```


:::

### Recover meta data from file names

Deliberate use of underscores (`_`) and hyphens (`-`) allows us to recover metadata from file names.

- Use underscores (`_`) to delimit units of metadata
- Use hyphens (`-`) to delimit words to improve human readability

::: {.panel-tabset}

## {{< fa brands r-project >}}

```{r filename="Code"}
flist <- list.files(pattern = "Plasmid") %>% head
stringr::str_split_fixed(flist, "[_\\.]", 5)
```

```{r filename="Output"}
#| code-copy: false
# date assay sampleset well
[1,] "2013-06-26" "BRAFWTNEGASSAY" "Plasmid-Cellline-100-1MutantFraction" "A01" "csv"
[2,] "2013-06-26" "BRAFWTNEGASSAY" "Plasmid-Cellline-100-1MutantFraction" "A02" "csv"
[3,] "2013-06-26" "BRAFWTNEGASSAY" "Plasmid-Cellline-100-1MutantFraction" "A03" "csv"
[4,] "2013-06-26" "BRAFWTNEGASSAY" "Plasmid-Cellline-100-1MutantFraction" "B01" "csv"
[5,] "2013-06-26" "BRAFWTNEGASSAY" "Plasmid-Cellline-100-1MutantFraction" "B02" "csv"
[6,] "2013-06-26" "BRAFWTNEGASSAY" "Plasmid-Cellline-100-1MutantFraction" "B03" "csv"
```


:::

### Summary

Names are more easily machine readable if it is

1. easy to search for files later
1. easy to narrow file lists based on names
1. easy to extract information from file names, for example by splitting

## Human readable

Names of files and folders should contain **information about its content**.
It should make it easy to figure out what something is based on the file name.
This idea connects to the concept of a slug from semantic urls.

::: {.callout-tip title="Find out more: What is a slug in a URL?" collapse="true"}

A slug is a part of a URL that is easy to read and understand.
It usually comes after the main website address and describes the content of the page in simple words.
Slugs are important for both humans and search engines because they give a clear idea of what the page is about.

For example:

- The full URL might be: `https://www.example.com/blog/how-to-cook-pasta`
- The slug in this URL is: `how-to-cook-pasta`

Good slugs are short, descriptive, and use hyphens to separate words.
This makes the URL more user-friendly and helps search engines understand the page content.

:::

Find the slug in the following file names:

```{zsh filename="Output"}
#| code-copy: false
01_marshal-data.r
02_pre-dea-filtering.r
03_dea-with-limma-voom.r
04_explore-dea-results.r
90_limma-model-term-name-fiasco.r
helper01_load-counts.r
helper02_load-exp-des.r
helper03_load-focus-statinf.r
helper04_extract-and-tidy.r
```

Which filenames make it easier to understand their content?

```{zsh filename="Output"}
#| code-copy: false
01_marshal-data.md
01_marshal-data.r
02_pre-dea-filtering.md
02_pre-dea-filtering.r
03_dea-with-limma-voom.md
03_dea-with-limma-voom.r
04_explore-dea-results.md
04_explore-dea-results.r
90_limma-model-term-name-fiasco.md
90_limma-model-term-name-fiasco.r
Makefile
figure
helper01_load-counts.r
helper02_load-exp-des.r
helper03_load-focus-statinf.r
helper04_extract-and-tidy.r
tmp.txt
```

```{zsh filename="Output"}
#| code-copy: false
01.md
01.r
02.md
02.r
03.md
03.r
04.md
04.r
90.md
90.r
Makefile
figure
helper01.r
helper02.r
helper03.r
helper04.r
tmp.txt
```

## Play well with default ordering

### Put something numeric first

#### Chronological order

```{zsh filename="Output"}
#| code-copy: false
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H01.csv
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H02.csv
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H03.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A01.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A02.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A03.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A04.csv
```

#### Logical order

```{zsh}
01_marshal-data.r
02_pre-dea-filtering.r
03_dea-with-limma-voom.r
04_explore-dea-results.r
90_limma-model-term-name-fiasco.r
helper01_load-counts.r
helper02_load-exp-des.r
helper03_load-focus-statinf.r
helper04_extract-and-tidy.r
```

#### Use the ISO 8601 standard for dates

`YYYY-MM-DD`

```{zsh filename="Output"}
#| code-copy: false
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H01.csv
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H02.csv
2013-06-26_BRAFWTNEGASSAY_Plasmid-Cellline-100-1MutantFraction_H03.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A01.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A02.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A03.csv
2014-02-17_BRAFWTNEGASSAY_FFPEDNA-CRC-1-41_A04.csv
```

```{r}
#| results: asis
#| eval: true
#| echo: false
cat(make_figure("xkcd-iso-8601"))
```

#### Consider the output:

Files that are later turned into HTML should be exclusively delimited by hyphens (`-`).

### Left-pad numbers with zeros

Left-padding of numbers refers to the practice of adding zeros to the left of a number to make it reach a certain length.
This can be useful, for example, when dealing with numerical data that needs to be formatted in a specific way for presentation or computational purposes.
By left-padding numbers, you can ensure that they all have the same length and are easier to compare or work with in a numerical context.

```{zsh filename="Output"}
#| code-copy: false
01_marshal-data.r
02_pre-dea-filtering.r
03_dea-with-limma-voom.r
04_explore-dea-results.r
90_limma-model-term-name-fiasco.r
helper01_load-counts.r
helper02_load-exp-des.r
helper03_load-focus-statinf.r
helper04_extract-and-tidy.r
```

If you don't left pad numbers, you get this:

```{zsh filename="Output"}
#| code-copy: false
10_final-figs-for-publication.R
1_data-cleaning.R
2_fit-model.R
```

## Summary

### Machine readable

1. File names are regular expression and globbing friendly
1. File names allow to recover meta data

### Human readable

Add scss code to disable the code-copy button on selected code chunks

1. File names contain info on content

### Plays well with default ordering

1. Put something numeric first
1. Use the ISO 8601 standard for dates: `YYYY-MM-DD`
1. Left pad other numbers with zeros

## References

@bryan2015

<https://speakerdeck.com/jennybc/how-to-name-files>
