---
title: "Research Data Management"
engine: knitr
execute:
  eval: false
  warning: false
  message: false
categories: [intermediate]
abstract: |
  In this chapter you will learn the basics of research data management.
---

## Introduction

::: {#vid-rdm-snafu}

{{< video https://www.youtube.com/watch?v=N2zK3sAtr-4?cc_load_policy=1&cc_lang_pref=en >}}

**"Data Sharing and Management Snafu in 3 Short Acts".**
The video by @hanson2019 from NYU Health Sciences Library features a cute animated panda researcher making a data sharing request to a bear researcher.
It does not go well.
The video aims to communicate some of the common and avoidable pitfalls of research data management in a light-hearted format.
Topics include storage, documentation, and file formats.
The video was produced using a tool called Xtranormal (now "nawmal") and the characters are strangely expressive while talking in robotic text-to-speech voices with a limited range of movement.
It was created and uploaded to YouTube by librarians at NYU School of Medicine in 2012 to be used for research data management education and has been included in numerous research data management courses and presentations.
License: Creative Commons Attribution 4.0 International ([CC BY 4.0](https://creativecommons.org/licenses/by/4.0/)).
Reused without modifications.

:::

See @vid-rdm-snafu ...

::: {.callout-warning title="The story of Reinhart & Rogoff"}
:::

:::{.callout-important title="Why you should stop using Excel!"}
:::

## Benefits of research data management

Managing data well throughout the research process is crucial for successful research outcomes, including:

- Upholding research integrity and reproducibility
- Improving research productivity
- Ensuring the accuracy, completeness, authenticity, and reliability of research data and records
- Saving time and resources in the future
- Strengthening data security and reducing the likelihood of data loss
- Avoiding redundant work by facilitating data sharing
- Meeting the grant requirements of funding bodies
- Adhering to industry and commerce practices.

## Motivation

### Modularity of research data

What we expect the data to look like ...

::: {layout-ncol=3}

```{r}
#| results: asis
#| eval: true
#| echo: false
cat(make_figure("remi-gau-data-source"))
```

```{r}
#| results: asis
#| eval: true
#| echo: false
cat(make_figure("remi-gau-data-raw"))
```

```{r}
#| results: asis
#| eval: true
#| echo: false
cat(make_figure("remi-gau-data-derivative"))
```

:::

What the dataset actually looks like ...

::: {layout-ncol=3}

```{r}
#| results: asis
#| eval: true
#| echo: false
cat(make_figure("remi-gau-data-shared"))
```

```{r}
#| results: asis
#| eval: true
#| echo: false
cat(make_figure("remi-gau-data-readme"))
```

:::

## FAIR principles

@wilkinson2016

```{r}
#| results: asis
#| eval: true
#| echo: false
cat(make_figure("nlm-nih-fair"))
```

## Data organization

### Data standards

```{r}
#| results: asis
#| eval: true
#| echo: false
cat(make_figure("xkcd-standards"))
```

Questions to ask yourself:

1. What data are you working with?
1. Which analysis pipelines are you working with?
1. **Is there a community standard?**

#### Example: Brain Imaging Data Structure (BIDS)

```{r}
#| results: asis
#| eval: true
#| echo: false
cat(make_figure("bids-dataset"))
```

@gorgolewski2016

#### Example: Psychological Data Structure (Psych-DS)

### Rectangular Data

If you do not have a community standard specifying the data organization, we highly recommend using the rules of **rectangular data** as proposed in @broman2018.

:::{.callout-tip title="Rules of rectangular data [@broman2018]"}
1. be consistent
1. choose good names for things (see Naming things)
1. write dates as YYYY-MM-DD (see Naming things)
1. no empty cells
1. put just one thing in a cell
1. make it a rectangle
1. create a data dictionary (see @sec-codebook)
1. no calculations in the raw data files
1. do not use font color or highlighting as data
1. make packups
1. use data validation to avoid errors
1. save the data in plain text files
:::

## Codebook / `.json` files {#sec-codebook}

## Acknowledgements

[Dr. Maike Kleemeyer](https://orcid.org/0000-0002-9388-5535) and the Research Data Management and Open Science working group at the [Max Planck Institute for Human Development, Berlin]({{< links.mpib >}}).

@gau2022

@gorgolewski2016
